Project for practical machine learning course
========================================================
The goal of your project is to predict the manner in which they did the exercise. 
This is the "classe" variable in the training set.
You may use any of the other
variables to predict with. You should create a report describing how you built your model,
how you used cross validation, what you think the expected out of
sample error is, and why you made the choices you did. You will also use your prediction model
to predict 20 different test cases.
Note : building model takes a long time due to kfold cross validation used and tuning the model parameters
 

```{r load test train csv files }
#   load training and test data data sets, these should be in default folder 
#   have a look at structure
#   classe is oucome variable, convert it factor variable
#   have a look at various variables and compare data structure of test and  training data sets

    train <- read.csv('pml-training.csv',stringsAsFactors =F); str(train); train <-train[-1]
    test <- read.csv('pml-testing.csv',stringsAsFactors =F); str(test); test <-test[-1]
    train$classe <- as.factor(train$classe)

```


```{r compare training and test data }

#  verify  training and test data structures match
#  only the last column does not match

    which( colnames(train)=="classe" )
    rnametr <-  names(train)
    rnamete <-names(test)
    rnametr ==   rnamete
```


```{r clean up data }
#  remove data columns  which are blank  or have all missing NA values
#  also get rid of person name and time of sampling data
    
    x1 <-train
    train <-x1[!sapply(x1, function(x)  { all(x == "" || is.na(x)) }   ) ]
    x2 <-test
    test <-x2[!sapply(x1, function(x)  { all(x == "" || is.na(x)) }   ) ]
    train <-train[,-(1:4)]; 
    test <- test[,-(1:4)]; 
```


```{r sample train data & split it for checking accuracy }



# sample only 2000 records from  training data set as original as has 19622 records
# also split data to trainValid and trainSample data sets    
    
    sampleRows = sample(1:dim(train)[1],size=2000,replace=F)
    trainSample = train[sampleRows,];row.names(trainSample)  <-NULL
    trainValid  = train[-sampleRows,];row.names(trainValid) <- NULL
```

 
 
```{r build models}

#  build models    
#  use k fold cross validation  (and the argument repeats controls the number of repetitions). K
# is controlled by the number argument and defaults to 10. Also parameters of random forest are tuned
# use random forest & caret package to build model and overall  accuracy was used as a measure of error
# note : it takes a long time to build the model & saved and loaded just in case
    library(caret) 
    ctrl <- trainControl(method = "repeatedcv",   repeats = 1,number =10)  
    modlRF <-train( as.factor(classe)  ~ .,trControl = ctrl,tuneLength = 5,  data = trainSample ,
                    preProc = c("center", "scale"),
                    method="rf")
                    

```


```{r validate the accurcy of model on unknown data set}


#   save model as it takes a long time to re-build it 
#   use the model to predict trainValid data set and compute accuracy
#   output shown below, accuaracy is about 97% which is acceptable

    save(modlRF, file = "modlRF.rda") 
    load(file = "modlRF.rda")
    modlRF 
    predValid  <-predict(modlRF, trainValid[-55])
    confusionMatrix(predValid, trainValid[,55],dnn = c("Prediction", "Reference"))

 
 
```


```{r predict for test data set}

#   use prediction model
#   to predict 20 different test cases. and save output as indivdual files
#   for checking on corsera platform for accuracy

    predTest <-predict(modlRF, test[-55])
     
    answers = rep("A", 20)
    pml_write_files = function(x){
        n = length(x)
        for(i in 1:n){
            filename = paste0("problem_id_",i,".txt")
            write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
        }
    }
    pml_write_files(predTest)
    predTest
    
 
```

